---
title: "Tibble Type"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tibble Type}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(canny.tudor)
```

Can you find the type of a `tibble::tibble`, meaning the types plural; that is,
one for each of the table columns? If you ask for `base::typeof` you only see
`"list"`.

## Take an example
Two columns: one of `int`, the other of `chr`.
```{r}
tibble::tibble(a = 1:10, b=letters[1:10])
```
You can access the column names using `base::names` and then apply the resulting
names to the table using the `[[` sub-setting operator, turning each column to
its underlying vector each in turn. Tibbles recommend using `[[` in newer code.

Finally `base:typeof` tells you the vector type.

## Sketch implementation
```{r}
types.of <- function(x)
  sapply(names(x), function(y) typeof(x[[y]]))
```
It answers a named vector of character describing the type of each column paired
with the name of the column.
```{r}
library(canny.tudor)
types.of(tibble::tibble(a=1:10, b=letters[1:10]))
```

## Data frames as well?
Does the same approach work on non-tibble data frames? It does because it only
relies on subset access operators.
```{r}
df <- data.frame(x = 1.1, y = I(letters[1:10]))
canny.tudor::types.of(df)
```

## Database frames
And finally, what about data frames lazily-loaded from a database? Do the
underlying column types manifest themselves? And do they appear _before_ or
_after_ collection of data by query execution? The simple answer is no, not
without first collecting some rows. This includes no rows, or limiting rows to
zero.

You might expect this result since the data frame does not yet exist, strictly
speaking; only the promise exists until some SQL runs.

### Try it
First create an SQLite database and set up a table, called `df` short for data
frame. Fill it with some test data.
```{r}
DBI::dbConnect(RSQLite::SQLite()) -> db
DBI::dbWriteTable(db, "df", data.frame(a=1:10, b=letters[1:10]))
dplyr::tbl(db, "df")
dplyr::tbl(db, "df") %>% types.of()
```
Above uses R 4.1+ built-in pipe operator; also uses `dplyr` to read back the
table lazily and derive the types. No joy. You cannot access lazy table columns
as you would an in-memory data frame.

Column subset answers `NULL` because the frame is not yet available locally.
```{r}
dplyr::tbl(db, "df")[["a"]]
sapply(dplyr::tbl(db, "df") %>% class(), \(x) methods(class = x))
```
However, you *can* effectively select no rows to derive an empty data frame
carrying empty column vectors but one empty vector for each column and with the
appropriate column type.
```{r}
dplyr::tbl(db, "df") %>%
  head(0) %>%
  dplyr::collect() %>%
  canny.tudor::types.of()
DBI::dbDisconnect(db)
```
Joy!
