---
title: "Tibble Type"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tibble Type}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(canny.tudor)
```

Can you find the type of a `tibble::tibble`, meaning the types plural; that is,
one for each of the table columns?

## Take an example
Two columns: one of `int`, the other of `chr`.
```{r}
tibble::tibble(a = 1:10, b=letters[1:10])
```
You can access the column names using `base::names` and then apply the resulting
names to the table using the `[[` sub-setting operator, turning each column to
its underlying vector each in turn. Tibbles recommend using `[[` in newer code.

Finally `base:typeof` tells you the vector type.

## Sketch implementation
```{r}
types.of <- function(x) 
  sapply(names(x), function(y) typeof(x[[y]]))
```
It answers a named vector of character describing the type of each column paired
with the name of the column.
```{r}
library(canny.tudor)
types.of(tibble::tibble(a=1:10, b=letters[1:10]))
```

## Data frames as well?
Does the same approach work on non-tibble data frames? It does because it only
relies on subset access operators.
```{r}
df <- data.frame(x = 1.1, y = I(letters[1:10]))
canny.tudor::types.of(df)
```

